Combinatorial interpretations of~holonomic recurrences

### 2021-09-15 article/main.tex
Pre-release

\usepackage{todonotes}
\setlength{\marginparwidth}{2.5cm}

@@

Generating functions are a way of describing infinite sequences using formal power series, where coefficients encode the consecutive elements. It allows us to incorporate many advanced methods of analysis of encoded objects.

Most recursive structures can be described using a~generating function equation, which can be then solved and transformed into a~holonomic recurrence, i.e.~recurrence that in order to define \(n\)-th element of the sequence requires just finite number of previous ones and coefficients that are polynomials in \(n\).

It allows us to quickly count objects of given structure with given size. At first glance, the recurrence seems to be closely related to the structure itself. It just seems like it should describe simple transformations resulting in constructing bigger objects from slightly smaller ones. In case of trees it could be for example adding nodes. However, there is no known method of interpreting it in the language of the combinatorial structures.

Up to today only a few interpretations of holonomic recurrences on trees have been found, including interpretation of holonomic recurrences for binary trees~\cite{binary}, unary-binary trees~\cite{motzkin}, and Schröder trees~\cite{schroder}. Unfortunately, the mentioned interpretations do not come with a method of finding them that could be generalized to more complex cases.

@@

Lambda terms are extremely popular structures used for analyzing functions and programming languages. Their simple recursive definition makes them very attractive. A better understanding of them could allow us to improve programming language compilers and in consequence improve every aspect of our lives. The interpretation of holonomic recurrence for trees of lambda terms itself could allow us to discover new ways of generating random lambda terms via extending smaller ones in similar way to Rémy's algorithm for sampling binary trees~\cite{remy,note}.

Throughout the thesis we try to accumulate known results and techniques that may help solving the problem of finding combinatorial interpretations. We also propose an~approach that helped us finding an~intermediate solution for the problem of finding combinatorial interpretation for holonomic recurrence of trees of lambda terms in the unary de Bruijn notation. We use the de Bruijn indices in order to get rid of letter-based variable names. Unary notation allows us to connect the size of the tree to the values of the de Bruijn indices.

@@

In Section~\ref{sec:approaches} we show approaches to the problem that helped to achieve the aforementioned results. All of them were tried in order to find the answer to the original problem of interpreting holonomic recurrence of trees of lambda terms in the unary de Bruijn notation. Unfortunately, none of them helped us enough to get the final result. Anyway, they are proven to be helpful when we are stuck, as most of them are able to divide problems into smaller parts and detect corner cases.

@@

\(\poly\) is a~set of all polynomials.

@@

\(f^{(n)}\) is implicitly defined as \(n\)-th derivative of function \(f\).

@@

Sibling of the node \(v\) is defined as the node \(u\) having the same parent as \(v\).

Uncle of the node \(v\) is defined as the~sibling of the parent of \(v\).

We say that node \(v\) is an~oneling \(\textiff\) \(v\) has no siblings. Note that the root is an oneling even if it has no parent.

We say that node \(v\) is a~twinling \(\textiff\) \(v\) has at least one sibling.

@@

Virtual node is an~artificially defined node that does not affect the structure definition (e.g.~we can define virtual leafs that are children of actual leaves of the binary tree. Such virtual leaves are not included when counting the number of children of the node, so the actual leaves are still called leaves).

@@

We say that node \(v\) is an~virtual-oneling \(\textiff\) \(v\) has no siblings being virtual nodes.

We say that node \(v\) is virtual-twinling \(\textiff\) \(v\) has at least one sibling being virtual node.

For a~binary tree, we say that node \(v\) is right (left) \(\textiff\) \(v\) is a~right (left) child of its parent.

For a~binary tree, we say that node \(v\) has a~right (left) parent \(u\) \(\textiff\) \(v\) is a~left (right) child of \(u\).

@@

Calligraphic \(\gf{X}\) is implicitly assumed to be a~generating function of the sequence \(X\) where \(X\) is any upper-case letter.

@@

Note that in order to compute the \(k\)-th element of the sequence, namely \(a_k\), we just need to find the~constant corresponding to indeterminate \(z^k\).

If we have a sequence counting some objects and a corresponding generating function, we say that this function counts the objects.

\begin{example}%
    \label{ex:bin_gf}
    Let's consider the recursive definition of binary trees shown in Figure~\ref{fig:binary_recursion}. We can create a~functional equation to count trees with exactly \(n\) internal nodes. To do so, we make internal nodes contribute to the generating function with \(z^1 = z\) and leaves with \(z^0 = 1\):

@@

\[\begin{array}{rclr}
        \gf{B}(z) &=& 1 + z\gf{B}^2(z)\\
                  &=& 1 + z\bigg(1 + z\gf{B}^2(z)\bigg)\bigg(1 + z\gf{B}^2(z)\bigg)\\
                  &=& 1 + \mhl{z}\bigg(1 + \mhl{z}\Big(\mhl{1} + z\gf{B}^2(z)\Big)\Big(\mhl{1} + z\gf{B}^2(z)\Big)\bigg)\bigg(\mhl{1} + z\gf{B}^2(z)\bigg) &(*)\\
                  &=& 1 + z\bigg(1 + z\Big(1 + z\gf{B}^2(z)\Big)\Big(1 + z\gf{B}^2(z)\Big)\bigg)\bigg(1 + z\Big(1 + z\gf{B}^2(z)\Big)\Big(1 + z\gf{B}^2(z)\Big)\bigg)\\
                  &=& 1 + z + 2 z^2 + \bigg(2 \gf{B}^2(z) + {\Big(1 + z \gf{B}^2(z)\Big)}^2\bigg) z^3
\end{array}\]

It is enough, as the term \(\bigg(2 \gf{B}^2(z) + {\Big(1 + z \gf{B}^2(z)\Big)}^2\bigg) z^3\) will contribute only to \(z^k\) where \(k > 2\). Therefore, we can see that there are \(2\) trees with exactly \(2\) internal nodes, as multiplier next to \(z^2\) is \(2\).

To visualize it even more, let's take a~look at the line marked with \((*)\). The highlighted terms describe following tree:

@@

\begin{example}
    The sequence \(a_n = \frac{5n - 3}{3n + 5}\) is holonomic, because it satisfies
    \[(3n + 5)(5n + 2) a_n - (5n - 3)(3n + 8) a_{n+1} = 0.\]
\end{example}

@@

The equivalence of these two definitions assures us that if we find a~generating function equation describing the desired objects satisfying Definition~\ref{def:holo_2}, we will be able to find holonomic recurrence mentioned in Definition~\ref{def:holo_1}~\cite{complexity}.

\begin{example}%
    \label{ex:gfun-rec}
    For generating function counting full binary trees defined in Example~\ref{ex:bin_gf}, i.e.:
    \[\gf{B}(z) = 1 + z\gf{B}^2(z)\]
    we can use GFUN~\cite{gfun} to immediately find a~holonomic recurrence of our interest:

@@

    The most interesting part for us is the relation among the finite number of consecutive elements of the sequence.
\end{example}

@@

Combinatorial interpretation is just a~fancy name for something you probably know and use. For example if you know the binomial coefficient \(\binom{n}{k}\), I bet it was introduced to you as something like: ``let's assume you have \(n\) distinguishable balls. If you want to choose \(k\) from them, then you can do it in \(\binom{n}{k}\) ways'' and now, whenever you see the term \(\binom{n}{k}\), you interpret it as the number of ways to choose \(k\) objects from the set of size \(n\). This is the combinatorial interpretation of the binomial coefficient.

@@

In the thesis we try to gather methods and results that may be helpful for anyone trying to find such an~interpretation, as we believe that it may give better insight into the nature of the structure and make it easier to study the structure's properties.

@@

In the thesis, if not stated otherwise, we will be working on rooted ordered unlabeled \(n\)-ary trees. To understand the terms, let's define them:
\begin{itemize}
    \item rooted tree is a~tree with single node marked as the root. This allows us to define a hierarchy of nodes using such terms as parent and child;
    \item ordered trees introduce order on children from the leftmost to the rightmost;
    \item unlabeled trees have nodes that are not distinguishable by their label. We will introduce soon kinds of nodes, but you shouldn't treat them as labels, but rather way to describe the structure of the node that is already present (e.g.~we can call node with just left child an~\(l\)~node, but it is not its label. If the right child was added to the \(l\)~node, it would not be an~\(l\)~node anymore).
\end{itemize}
Additionally we allow that an \(n\)-ary node has some of its children unset (e.g.~we can define a~binary node with just the left child set, but not the right one).

@@

\begin{definition}
    For a single node \(v\), the pre-order notation is defined as \(\enc{v}\).
\end{definition}

Here, \(\enc{v}\) is defined as follows:
\begin{itemize}
    \item if \(v\) is an~existing node, \(\enc{v} = \n\);
    \item if \(v\) is a~non-existing node (i.e.~we describe non-existing child of some node), \(\enc{v} = \no\).
\end{itemize}

@@

You can note that the encoding of a~given tree is unique. Unfortunately, not every sequence encodes a proper tree.

@@

\begin{definition}
    Let \(v\) be a~node described using the \(k\)-th symbol of\;\(\enc{T}\). The prefix of \(T\) in relation to \(v\), namely \(\tPre(T, v)\), is defined using pre-order notation as a~prefix of \enc{T} of size \((k - 1)\).
\end{definition}

\begin{definition}
    Let \(n\) be the length of the sequence \(\enc{T}\). The suffix of \(T\) in relation to \(v\), namely \(\enc{\tSuf(T, v)}\), is defined using pre-order notation as a~suffix of \enc{T} of size \(n - \size{\enc{\tPre(T, v)}} - \size{\enc{\tSub(v)}}\). Note that the suffix may be an ordered forest.
\end{definition}

In other words, the prefix of the tree \(T\) in relation to \(v\) describes a~part of the tree that appears before the subtree rooted in \(v\) in pre-order notation. Suffix describes a part of the tree that appears after it.

@@

Throughout the thesis we define some operations using this notation. Valid pre-order notation of an \(n\)-ary tree (note that not all sequences are valid pre-order notations) can be easily translated back into a~tree. In order to do so, we just take the first symbol of the notation and create the root accordingly. Then for each of the \(n\) children of the root, in sequence, we decode it recursively as if it was a whole tree. Each such operation consumes some prefix of the notation. At the end we should be left with no symbols in the sequence.

As \(T\) and \(\enc{T}\) describe the same structures, notation of trees \(T\) and its encoding \(\enc{T}\) will be used interchangeably.

@@

In this section we show combinatorial interpretations of holonomic recurrences for binary trees, unary-binary trees and Schröder trees. As to our knowledge these are the only tree classes that have commonly known combinatorial interpretations. Also the thought process leading to these interpretations is not obvious.

Each of these interpretations has its unique character, like a different selection of the size definition or a different approach to arranging the terms of the holonomic recurrence.

These interpretations have been a source of ideas that we were incorporating in order to make our partial interpretation for the case of trees of lambda terms in unary de Bruijn notation.

@@

Let's start with probably the most known interpretation, full binary trees~\cite{binary}.

@@

\[\gf{B}(z) = 1 + z\gf{B}^2(z)\]

@@

This equation actually have \(2\) solutions, but we can eliminate one of them due to assumption that we are counting objects, i.e.~we are interested only in solutions that give us natural numbers.

@@

It is easy to see that a~tree of size \(n\) (i.e.~with \(n\) internal nodes) has \(n + 1\) leaves and therefore \(2n + 1\) nodes in total. Analogically the tree of size \(n + 1\) has \(n + 2\) leaves and \(2n + 3\) nodes in total. If you see some resemblance to our holonomic recurrence, you are perfectly right.

@@

At the moment the left side of the recurrence is related to the trees of size \(n + 1\), the right, on the other hand, of size \(n\). Therefore, we need to apply some transformations on the trees.

Let's define two local operations for a tree \(T\) with a~single node \(v\) pointed: \(R_l(T, v)\) and \(R_r(T, v)\). The term \textit{local} means that the operation can operate on just some finite fragment of a~tree with predefined size (e.g.~node and its 2 children, but cannot traverse the whole tree).

\begin{definition}
    \(\enc{R_l(T, v)} = \enc{\tPre(T, v)} \n \pointed{\n} \no \no \enc{\tSub(v)} \enc{\tSuf(T, v)}\)
\end{definition}

@@

\begin{definition}
    \(\enc{R_r(T, v)} = \enc{\tPre(T, v)} \n \enc{\tSub(v)} \pointed{\n} \no \no \enc{\tSuf(T, v)} \)
\end{definition}

@@

We can now take the bijective interpretation of \((2n + 1) B(n)\), i.e.~trees of size \(n\) with a~single node pointed. For each of such trees, namely \(T\), we create a new one by applying \(R_l\) and \(R_r\) to the pair \(T', v\), where \(T'\) is \(T\) with removed pointers and \(v\) is a~node that is pointed in \(T\). Therefore we generate all trees from \(2(2n + 1) B(n)\).
\todo{rewrite}

The only thing that's left for us to verify is if we generate all the trees of size \(n + 1\) with pointed leaves. As from the recursion, we know that \(2 (2n + 1) B(n)\) describe trees with a single leaf pointed. It is enough to check if we do not generate any duplicate using \(R_l\) and \(R_r\). It can happen in only 2 ways:
\begin{enumerate}
    \item operation itself generates duplicates or
    \item 2 different operations generate the same tree.
\end{enumerate}

First case can also be split into 2 subcases. Without loss of generality we can analyze just one of the operations, let it be \(R_l\). The second case is symmetric.
\begin{itemize}
    \item for some nodes \(u \neq v\), \(R_l(T, u) = R_l(T, v)\). Let \(i_u, i_v\) be the indices of respectively \(\enc{u}\) and \(\enc{v}\) in \(\enc{T}\). Let \(\pointed{i}_u, \pointed{i}_v\) be the indices of pointed node respectively in \(\enc{R_l(T, u)}\) and \(\enc{R_l(T, v)}\). Due to the definition of \(R_l\), we get that \(\pointed{i}_u = i_u + 1\) and \(\pointed{i}_v = i_v + 1\), which implies that \(u = v\) what we assumed false.
    \item for some trees \(T \neq S\), \(R_l(T, u) = R_l(S, v)\). Let \(i_u, i_v\) be the indices of respectively \(\enc{u}\) in \(\enc{T}\) and \(\enc{v}\) in \(\enc{S}\). Let \(\pointed{i}_u, \pointed{i}_v\) be the indices of pointed node respectively in \(\enc{R_l(T, u)}\) and \(\enc{R_l(S, v)}\). If \(\pointed{i}_u \neq \pointed{i}_v\), this case is trivially false, so we can assume that \(\pointed{i}_u = \pointed{i}_v\). Therefore, again from the definition of \(R_l\), we have that \(i_u = i_v\). That implies that \(\enc{\tPre(T, u)} \enc{\tSub(u)} \enc{\tSuf(T, u)} = \enc{\tPre(S, v)} \enc{\tSub(v)} \enc{\tSuf(S, v)} \), which then sums up to \(\enc{T} = \enc{S}\), which we assumed false.
\end{itemize}

In the second case it is enough to notice that \(R_l\) generates trees with pointed nodes being a~left child, and \(R_r\) generates trees with pointed nodes being a right child. That immediately provides us with the conclusion that we are generating no duplicates.

@@

\begin{definition}%
    \label{def:unary_binary_tree}
    An unary-binary tree is an~ordered tree in which any node can have either 0, 1 or 2 children.
\end{definition}

We do not allow nodes with gaps here, e.g.~a~binary node with just one child or unary node with no children.

We will define the size as a number of edges. Note that a~tree having \(n\) edges has exactly \(n + 1\) nodes.

We can easily derive the generating function:

\[\gf{U}(z) = 1 + z\gf{U}(z) + z^2 \gf{U}^2(z)\]

And with a~bit of Maple magic~\cite{gfun}, we immediately get the following recurrence:

\[(n + 2)u_{n} = (2n + 1)u_{n - 1} + 3(n - 1)u_{n - 2}\]

@@

\begin{definition}%
    \label{def:unary_binary_2}
    A left-leaning binary tree is a~binary tree with the following constraint added: if a~node has a~right child, it also has a~left child. Of course a node with just a left child can appear in such tree.
\end{definition}

The bijection between unary-binary trees and left-leaning binary ones is very simple. We just represent unary nodes as nodes with just a~left child. The other cases translate immediately to themselves. Please note that this transformation preserves the size of the tree (i.e.~number of edges).

@@

\begin{definition}
    Virtual node is an~artificially defined node that does not affect the structure definition (i.e.~they are not included when counting the number of children of the node).
\end{definition}

\begin{definition}
    Actual node is a~node that is not virtual.
\end{definition}

@@

We will be using \(U(n)\) to describe sets of trees corresponding to \(u_n\) from the holonomic recurrence. The set \((n + 2) U(n)\) can be interpreted as trees of size \(n\) with a single virtual leaf pointed.

@@

\begin{interpretation}
To present the interpretation in bijective language, we will have to introduce several types of the nodes.

\begin{enumerate}
    \item Vro --- virtual leaf being right virtual-oneling;
    \item Vrt --- virtual leaf being right virtual-twinling. We will divide this case into:
        \begin{enumerate}
            \item VrtUrv --- Vrt which uncle is a~right virtual leaf;
            \item VrtUra --- Vrt which uncle is a~right actual node;
            \item VrtUla --- Vrt which uncle is a~left actual node.
        \end{enumerate}
    \item Vlt --- virtual leaf left virtual-twinling. We can divide also this case into:
        \begin{enumerate}
            \item VltUrv --- Vlt which uncle is a~right virtual leaf;
            \item VltUra --- Vlt which uncle is a~right actual node;
            \item VltUla --- Vlt which uncle is a~left actual node;
        \end{enumerate}
    \item A --- actual node.
\end{enumerate}

@@

All we need to show are transformations from the smaller cases, i.e.~pointed trees from \(U(n-1)\) and \(U(n-2)\) to \(U(n)\).

@@

In the case of \(U(n-2)\), the multiplier is \(3(n-1)\), which suggests that we should perform \(3\) distinct types of transformations on the trees with a~pointed actual node.

In the table below we show transformations that are applied to trees from \(U(n-1)\) and \(U(n-2)\) in order to achieve trees from \(U(n)\). We mark pointed node with \(\pointed{}\). You can find the case name in the bottom left corners of the cells.

@@

Let's represent node \(v\) with a~weight \(n\) as \(v\weighted{n}\).

>>>>

\usepackage[disable]{todonotes}
\setlength{\marginparwidth}{2.5cm}

@@

Generating functions are a way of describing infinite sequences using formal power series, whose coefficients encode the consecutive elements. They allow us to incorporate many advanced methods of analysis of encoded objects.

Many recursive structures can be described using a~generating function equation, which can be then solved and transformed into a~holonomic recurrence, i.e.~a~recurrence that in order to define the \(n\)-th element of the sequence requires just a finite number of previous ones and coefficients that are polynomials in \(n\).

Holonomic recurrences allow us to quickly count objects of a given structure of a given size. At the first glance, the recurrence seems to be closely related to the associated structure. It just seems like it should describe some simple transformations resulting in constructing bigger objects from slightly smaller ones. In case of trees it could be for example adding nodes. However, there is no known method of interpreting the holonomic recurrence in the language of the combinatorial structures.

Up to today only a few interpretations of holonomic recurrences on trees have been found, including interpretations of holonomic recurrences for binary trees~\cite{binary}, unary-binary trees~\cite{motzkin}, and Schröder trees~\cite{schroder}. Unfortunately, the mentioned interpretations do not come with a method of finding them that could be generalized to more complex cases.

@@

Lambda terms are extremely popular structures used for analyzing functions and programming languages. Their simple recursive definition makes them very attractive. A better understanding of them could allow us to improve programming language compilers and in consequence improve every aspect of our lives. The interpretation of a holonomic recurrence for trees of lambda terms itself could allow us to discover new ways of generating random lambda terms via extending smaller ones in a similar way to Rémy's algorithm for sampling binary trees~\cite{remy,note}.

Throughout the thesis we try to accumulate known results and techniques that may help solving the problem of finding combinatorial interpretations. We also propose an~approach that helped us to provide an~intermediate solution for the problem of finding combinatorial interpretation for a holonomic recurrence of trees of lambda terms in the unary de Bruijn notation. We use the de Bruijn indices in order to get rid of letter-based variable names. The unary notation allows us to connect the size of a tree to the values of the de Bruijn indices.

@@

In Section~\ref{sec:approaches} we show approaches to the problem that helped us to achieve the aforementioned results. All of them were tried in order to find an answer to the original problem of interpreting the holonomic recurrence of trees of lambda terms in the unary de Bruijn notation. Unfortunately, none of them helped us enough to get the final result. Anyway, they are proven to be helpful when we are stuck, as most of them are able to divide problems into smaller parts and detect corner cases.

@@

\(\poly\) is a~set of all polynomials.\todo{Fix poly definition}

@@

\(f^{(n)}\) is implicitly defined as the \(n\)-th derivative of a function \(f\).

@@

Sibling of a node \(v\) is defined as a node \(u\) having the same parent as \(v\).

Uncle of a node \(v\) is defined as a~sibling of the parent of \(v\).

We say that a node \(v\) is an~oneling \(\textiff\) \(v\) has no siblings. Note that the root is an oneling even if it has no parent.

We say that a node \(v\) is a~twinling \(\textiff\) \(v\) has at least one sibling.

@@

Virtual node is an~artificially defined node that does not affect the definition of a structure (e.g.~we can define virtual leafs that are children of actual leaves of a binary tree. Such virtual leaves are not included when counting the number of children of the node, so the actual leaves are still called leaves).

@@

We say that a node \(v\) is a~virtual-oneling \(\textiff\) \(v\) has no siblings being a virtual nodes.

We say that a node \(v\) is a~virtual-twinling \(\textiff\) \(v\) has at least one sibling being a virtual node.

For a~binary tree, we say that a node \(v\) is right (left) \(\textiff\) \(v\) is a~right (left) child of its parent.

For a~binary tree, we say that a node \(v\) has a~right (left) parent \(u\) \(\textiff\) \(v\) is a~left (right) child of \(u\).

@@

Calligraphic \(\gf{X}\) is implicitly assumed to be a~generating function of the sequence \(X\), where \(X\) is any upper-case letter.

@@

Note that in order to compute the \(k\)-th element of the sequence, namely \(a_k\), we just need to find the~constant corresponding to the indeterminate \(z^k\).

If we have a sequence counting some objects and the corresponding generating function, we say that this function counts the objects.

\begin{example}%
    \label{ex:bin_gf}
    Let's consider the recursive definition of binary trees shown in Figure~\ref{fig:binary_recursion}. We can construct a~functional equation involving the generating function counting trees with exactly \(n\) internal nodes. To do so, we make internal nodes contribute to the generating function with \(z^1 = z\) and leaves with \(z^0 = 1\):

@@

\[\begin{array}{rclr}
        \gf{B}(z) &=& 1 + z\gf{B}^2(z)\\
                  &=& 1 + z\bigg(1 + z\gf{B}^2(z)\bigg)\bigg(1 + z\gf{B}^2(z)\bigg)\\
                  &=& 1 + \mhl{z}\bigg(1 + \mhl{z}\Big(\mhl{1} + z\gf{B}^2(z)\Big)\Big(\mhl{1} + z\gf{B}^2(z)\Big)\bigg)\bigg(\mhl{1} + z\gf{B}^2(z)\bigg) &(*)\\
                  &=& 1 + z\bigg(1 + z\Big(1 + z\gf{B}^2(z)\Big)\Big(1 + z\gf{B}^2(z)\Big)\bigg)\bigg(1 + z\Big(1 + z\gf{B}^2(z)\Big)\Big(1 + z\gf{B}^2(z)\Big)\bigg)\\
                  &=& 1 + z + 2 z^2 + \bigg(2 \gf{B}^2(z) + {\Big(1 + z \gf{B}^2(z)\Big)}^2\bigg) z^3.
\end{array}\]

It is enough, as the term \(\bigg(2 \gf{B}^2(z) + {\Big(1 + z \gf{B}^2(z)\Big)}^2\bigg) z^3\) will contribute only to \(z^k\) where \(k > 2\). Therefore, we can see that there are \(2\) trees with exactly \(2\) internal nodes, as the multiplier next to \(z^2\) is \(2\).

To visualize it even better, let's take a~look at the line marked with \((*)\). The highlighted terms describe the following tree:

@@

\begin{example}
    The sequence \(a_n = \frac{5n - 3}{3n + 5}\) is holonomic, because it satisfies the following equation:
    \[(3n + 5)(5n + 2) a_n - (5n - 3)(3n + 8) a_{n+1} = 0.\]
\end{example}

@@

The equivalence of these two definitions assures us that if we find a~generating function equation describing the desired objects satisfying Definition~\ref{def:holo_2}, we will be able to find a holonomic recurrence mentioned in Definition~\ref{def:holo_1}~\cite{complexity}.

\begin{example}%
    \label{ex:gfun-rec}
    For the generating function counting full binary trees defined in Example~\ref{ex:bin_gf}, i.e.:
    \[\gf{B}(z) = 1 + z\gf{B}^2(z),\]
    we can use GFUN~\cite{gfun} to immediately find a~holonomic recurrence of our interest:

@@

    The most interesting part for us is the relation among the finite number of consecutive elements of the sequence \((b_n)_{n \in \N}\).
\end{example}

@@

Combinatorial interpretation is just a~fancy name for something you probably know and use. For example if you know the binomial coefficient \(\binom{n}{k}\), we can bet it was introduced to you as something like: ``let's assume you have \(n\) distinguishable balls. If you want to choose \(k\) from them, then you can do it in \(\binom{n}{k}\) ways'' and now, whenever you see the term \(\binom{n}{k}\), you interpret it as the number of ways to choose \(k\) objects from the set of size \(n\). This is a combinatorial interpretation of the binomial coefficient.

@@

In the thesis we try to gather methods and results that may be helpful for anyone trying to find such an~interpretation, as we believe that it may give a better insight into the nature of the structure and make it easier to study structure's properties.

@@

In the thesis, if not stated otherwise, we will be working with rooted ordered unlabeled \(n\)-ary trees. To understand the terms, let's define them:
\begin{itemize}
    \item rooted tree is a~tree with a single node marked as the root. This allows us to define a hierarchy of nodes using such terms as parent and child;
    \item ordered trees introduce an order on children from the leftmost to the rightmost;
    \item unlabeled trees have nodes that do not have any labels we could use to distinguish them. We will introduce soon kinds of nodes, but you shouldn't treat them as labels, but rather a way to describe the structure of a node that is already present (e.g.~we can call a node with just a left child an~\(l\)~node, but it is not its label. If the right child was added to the \(l\)~node, it would not be an~\(l\)~node anymore).
\end{itemize}
Additionally we allow an \(n\)-ary node to have some of its children unset (e.g.~we can define a~binary node with just the left child set, but not the right one).

@@

\begin{definition}
    For a single node \(v\), the pre-order notation is defined as \(\enc{v}\), where:

    \begin{itemize}
        \item if \(v\) is an~existing node, \(\enc{v} = \n\);
        \item if \(v\) is a~non-existing node (i.e.~we describe non-existing child of some node), \(\enc{v} = \no\).
    \end{itemize}
\end{definition}

@@

You can note that the encoding of a~given tree is unique. Unfortunately, not every sequence encodes a proper tree.
\todo{example}

@@

\begin{definition}
    Let \(v\) be a~node described using the \(k\)-th symbol of\;\(\enc{T}\). The prefix of \(T\) in relation to \(v\), namely \(\tPre(T, v)\), is defined using the pre-order notation as a~prefix of \enc{T} of size \((k - 1)\).
\end{definition}

\begin{definition}
    Let \(n\) be the length of the sequence \(\enc{T}\). The suffix of \(T\) in relation to \(v\), namely \(\enc{\tSuf(T, v)}\), is defined using the pre-order notation as a~suffix of \enc{T} of size \(n - \size{\enc{\tPre(T, v)}} - \size{\enc{\tSub(v)}}\). Note that the suffix is an ordered forest (not necessarily a single tree).
\end{definition}

In other words, the prefix of a tree \(T\) in relation to \(v\) describes a~part of the tree that appears before the subtree rooted in \(v\) in the pre-order notation. The suffix describes a part of the tree that appears after it.

@@

Throughout the thesis we define some operations using this notation. A valid pre-order notation of an \(n\)-ary tree (note that not all sequences are valid pre-order notations) can be easily translated back into a~tree. In order to do so, we just take the first symbol of the notation and create the root accordingly. Then for each of the \(n\) children of the root, in sequence, we decode it recursively as if it was a whole tree. Each such operation consumes some prefix of the notation. At the end we should be left with no symbols in the sequence.

As \(T\) and \(\enc{T}\) describe the same structures, notation of a tree \(T\) and its encoding \(\enc{T}\) will be used interchangeably.

@@

In this section we show combinatorial interpretations of holonomic recurrences for binary trees, unary-binary trees and Schröder trees. As to our knowledge these are the only tree classes that have commonly known combinatorial interpretations. Moreover, the thought process leading to these interpretations is not obvious.

Each of these interpretations has its unique character, like a different selection of the size definition or a different approach to arranging the terms of a holonomic recurrence.

These interpretations have been a source of ideas that we were incorporating in order to make our partial interpretation for the case of trees of lambda terms in the unary de Bruijn notation.

@@

Let's start with probably the most known interpretation, namely for full binary trees~\cite{binary}.

@@

@@

That equation has actually \(2\) solutions, but we can eliminate one of them due to assumption that we are counting objects, i.e.~we are interested only in solutions that give us natural numbers.

@@

It is easy to see that a~tree of size \(n\) (i.e.~with \(n\) internal nodes) has \(n + 1\) leaves and therefore \(2n + 1\) nodes in total. Analogically a tree of size \(n + 1\) has \(n + 2\) leaves and \(2n + 3\) nodes in total. If you see some resemblance to our holonomic recurrence, you are perfectly right.

@@

At the moment the left side of the recurrence is related to trees of size \(n + 1\), the right, on the other hand, of size \(n\). Therefore, we need to apply some transformations on the trees.

Let's define two local operations on a tree \(T\) with a~single node \(v\) pointed: \(R_l(T, v)\) and \(R_r(T, v)\). The term \textit{local} means that the operation can operate on just some finite fragment of a~tree with predefined size (e.g.~a~node and its 2 children, but cannot traverse the whole tree).

\begin{definition}
    \(\enc{R_l(T, v)} := \enc{\tPre(T, v)} \n \pointed{\n} \no \no \enc{\tSub(v)} \enc{\tSuf(T, v)}\)
\end{definition}

@@

\begin{definition}
    \(\enc{R_r(T, v)} := \enc{\tPre(T, v)} \n \enc{\tSub(v)} \pointed{\n} \no \no \enc{\tSuf(T, v)} \)
\end{definition}

@@

We can now take the bijective interpretation of \((2n + 1) B(n)\), i.e.~trees of size \(n\) with a~single node pointed. For each of such trees, namely \(T\), we create a new one by applying \(R_l\) and \(R_r\) to the pair \((T', v)\), where \(T'\) is \(T\) with removed pointers and \(v\) is a~node that is pointed in \(T\). Therefore we generate all trees from \(2(2n + 1) B(n)\).
\todo{rewrite}

The only thing that is left for us is to verify if we generate all the trees of size \(n + 1\) with pointed leaves. As from the recursion, we know that \(2 (2n + 1) B(n)\) describes trees with a single leaf pointed. It is enough to check if we do not generate any duplicates using \(R_l\) and \(R_r\). It can happen in only 2 ways:
\begin{enumerate}
    \item operation itself generates duplicates or
    \item 2 different operations generate the same tree.
\end{enumerate}

The first case can also be split into 2 subcases. Without loss of generality we can analyze just one of the operations, let it be \(R_l\). The second case is symmetric.
\begin{itemize}
    \item For some distinct nodes \(u\) and \(v\), \(R_l(T, u) = R_l(T, v)\). Let \(i_u, i_v\) be the indices of respectively \(\enc{u}\) and \(\enc{v}\) in \(\enc{T}\). Let \(\pointed{i}_u, \pointed{i}_v\) be the indices of a pointed node respectively in \(\enc{R_l(T, u)}\) and \(\enc{R_l(T, v)}\). Due to the definition of \(R_l\), we get that \(\pointed{i}_u = i_u + 1\) and \(\pointed{i}_v = i_v + 1\), which implies that \(u = v\) what we assumed false.
    \item For some trees \(T \neq S\), \(R_l(T, u) = R_l(S, v)\). Let \(i_u, i_v\) be the indices of respectively \(\enc{u}\) in \(\enc{T}\) and \(\enc{v}\) in \(\enc{S}\). Let \(\pointed{i}_u, \pointed{i}_v\) be the indices of a pointed node respectively in \(\enc{R_l(T, u)}\) and \(\enc{R_l(S, v)}\). If \(\pointed{i}_u \neq \pointed{i}_v\), this case is trivially false, so we can assume that \(\pointed{i}_u = \pointed{i}_v\). Therefore, again from the definition of \(R_l\), we have that \(i_u = i_v\). That implies that \(\enc{\tPre(T, u)} \enc{\tSub(u)} \enc{\tSuf(T, u)} = \enc{\tPre(S, v)} \enc{\tSub(v)} \enc{\tSuf(S, v)} \), which then sums up to \(\enc{T} = \enc{S}\), which we assumed false.
\end{itemize}

In the second case it is enough to notice that \(R_l\) generates trees with pointed nodes being a~left child, and \(R_r\) generates trees with pointed nodes being a right child. This immediately provides us with the conclusion that we are generating no duplicates.

@@

\begin{definition}%
    \label{def:unary_binary_tree}
    A unary-binary tree is an~ordered tree in which any node can have either 0, 1 or 2 children.
\end{definition}

We do not allow nodes with gaps here, e.g.~a~binary node with just one child or a unary node with no children.

We will define the size of a tree as the number of the edges. Note that a~tree having \(n\) edges has exactly \(n + 1\) nodes.

We can easily derive the generating function equation:

\[\gf{U}(z) = 1 + z\gf{U}(z) + z^2 \gf{U}^2(z).\]

With a~bit of Maple magic~\cite{gfun}, we immediately get the following recurrence:

@@

\begin{definition}%
    \label{def:unary_binary_2}
    A left-leaning binary tree is a~binary tree with the following constraint added: if a~node has a~right child, it also has a~left child. Of course a node with just a left child can appear in such a tree.
\end{definition}

The bijection between unary-binary trees and left-leaning binary ones is very simple. We just represent unary nodes as nodes with just a~left child. The other cases translate immediately to themselves. Please note that this transformation preserves the size of a tree (i.e.~the~number of edges).

@@

\begin{definition}
    A virtual node is an~artificially defined node that does not affect the structure definition (i.e.~it is not included when counting the number of children of the node).
\end{definition}

\begin{definition}
    An actual node is a~node that is not virtual.
\end{definition}

@@

We will be using \(U(n)\) to describe the set of trees corresponding to \(u_n\) from the holonomic recurrence. The set \((n + 2) U(n)\) can be interpreted as the set of trees of size \(n\) with a single virtual leaf pointed.

@@

\begin{interpretation}
To present the interpretation in the bijective language, we will have to introduce several types of nodes.

\begin{enumerate}
    \item Vro --- virtual leaf being a~right virtual-oneling;
    \item Vrt --- virtual leaf being a~right virtual-twinling. We will divide this case into:
        \begin{enumerate}
            \item VrtUrv --- Vrt which uncle is a~right virtual leaf;
            \item VrtUra --- Vrt which uncle is a~right actual node;
            \item VrtUla --- Vrt which uncle is a~left actual node.
        \end{enumerate}
    \item Vlt --- virtual leaf being a~left virtual-twinling. We can divide also this case into:
        \begin{enumerate}
            \item VltUrv --- Vlt which uncle is a~right virtual leaf;
            \item VltUra --- Vlt which uncle is a~right actual node;
            \item VltUla --- Vlt which uncle is a~left actual node;
        \end{enumerate}
    \item A --- actual node.
\end{enumerate}

@@

All we need to show are transformations from smaller cases, i.e.~pointed trees from \(U(n-1)\) and \(U(n-2)\) to \(U(n)\).

@@

In the case of \(U(n-2)\), the multiplier is \(3(n-1)\), which suggests that we should perform \(3\) distinct types of transformations on trees with a~pointed actual node.

In the table below we show transformations that are applied to trees from \(U(n-1)\) and \(U(n-2)\) in order to achieve trees from \(U(n)\). We mark a pointed node with \(\pointed{}\). You can find the case name in the bottom left corner of each cell.

@@

Let's represent node \(v\) with a~weight \(w\) as \(v\weighted{w}\).
